---
title: "Supplement A: Modeling Radiocarbon Dates"
date: today
published-title: "COMPILED"
format:
  html: 
    embed-resources: true
    fig-align: center
    fig-dpi: 300
    fig-responsive: true
    number-sections: true
    number-depth: 3
    toc: true
    toc-depth: 3
    smooth-scroll: true
    theme: cosmo
    highlight-style: a11y
    code-tools: true
    code-links:
      - text: kbvernon/netc14
        icon: github
        href: https://github.com/kbvernon/netc14
execute: 
  echo: true
  warning: false
  error: false
---

```{r}
#| label: time-code-chunks
#| echo: false
# we are deep into overkill territory here
knitr::knit_hooks$set(
  time_it = local({
    function(before, options) {
      if (before) {
        assign('.start_time', Sys.time(), envir = .GlobalEnv)
      } else {
        proc_time <- difftime(Sys.time(), .start_time, units = "secs")
        proc_time <- round(as.numeric(proc_time), 2)

        mins <- proc_time %/% 60
        secs <- proc_time %% 60

        proc_time <- if (mins > 0) {
          sprintf("%dm %.1fs", mins, secs)
        } else {
          sprintf("%.1fs", secs)
        }

        sprintf(
          "<p class='chunk-timing'>%s</p>",
          paste0("Code execution time: ", proc_time)
        )
      }
    }
  })
)
```

```{=html}
<style>
.code-copy-outer-scaffold:has(.chunk-timing) {
  position: relative;
}

.code-copy-outer-scaffold:has(.chunk-timing) .sourceCode.cell-code {
  margin-bottom: 2em;
}

.chunk-timing {
  position: absolute;
  bottom: -1.5em;
  right: 0;
  text-align: right;
  font-size: 0.75em;
  margin: 0;
  padding: 0;
  color: rgba(150, 150, 150, 1);
}

p:has(img) { 
  text-align: center; 
}
</style>

<script>
document.addEventListener("DOMContentLoaded", function() { 
  document.querySelectorAll(".chunk-timing").forEach(el => {
    const code = el.closest(".cell").querySelector(".code-copy-outer-scaffold");
    if (code) code.appendChild(el);    
  });
});  
</script>
```

## Introduction

**Goal:** Train a neural network to calibrate and summarize radiocarbon samples.

**Data:** For model evaluation, data are simulated according to the generative 
model. For comparative purposes, we also apply models to four real world 
datasets.

**Method:** Simulations are used to evaluate different modeling techniques,
allowing us to compare estimates against true and known ages and densities.

## R Preamble

```{r}
#| label: library
library(carbondate)
library(parallel)
library(svglite)
library(torch)
```

**Generic Methods**

In this analysis, we rely heavily on R's S3 system of generics and methods. This
allows us to simplify the API, making it more interpretable, but more 
importantly, it makes programming simulations much, much easier. Crucially, we
rely on four generic methods provided by base R: `print()`, `plot()`, 
`predict()`, and `density()`. The `predict()` method is implemented for models 
to estimate the true age of each radiocarbon sample. The `density()` method is 
implemented for models to estimate the true probability distribution. The print 
methods are mainly there for inspecting the models interactively. The plot 
methods help ensure consistent theming for figures.

```{r}
#| label: source-functions
#| output: false
fns <- c(
  "bgmm.R",
  "calibrate.R",
  "ckde.R",
  "constants.R",
  "dataframe-helpers.R",
  "dpmm.R",
  "metrics.R",
  "mixture-helpers.R",
  "netc14.R",
  "plot-calibration.R",
  "plot-density.R",
  "plot-helpers.R",
  "plot-theme.R",
  "print.R",
  "sample-dates.R",
  "simulate-chronology.R",
  "spd.R"
)

lapply(file.path("R", fns), source, verbose = FALSE)
```

**Constants**

```{r}
#| label: constants
#| file: R/constants.R
#| eval: false
```

## Data

### Load Calibration

For this modeling exercise, we use the most recent version of IntCal or 
IntCal20. Note that we convert means and errors to fraction modern (`f14c_mu` 
and `f14c_sigma`) since it is more reliable for older samples that may have 
asymmetric uncertainty around the mean.

```{r}
#| label: intcal20
intcal20 <- structure(
  read.csv("data/intcal20.csv"),
  class = c("Calibration", "data.frame"),
  curve = "IntCal20"
)

intcal20 <- transform(intcal20, f14c_mu = exp(-c14_age / LIBBY))
intcal20 <- transform(intcal20, f14c_sigma = f14c_mu * c14_error / LIBBY)

intcal20
```

### Simulate Chronology

We refer to the underlying probability density or age distribution as a 
"chronology." To simulate a chronology, we randomly sample mixture parameters 
from various distributions: Gamma for the weights and standard deviations, 
Discrete Uniform for the means. We then evaluate the mixture PDF over the entire 
interval at an annual resolution and then normalize those probabilities to sum 
to one, in effect generating a truncated and discrete empirical distribution. 

```{r}
#| label: function-simulate-chronology
#| file: R/simulate-chronology.R
#| code-fold: true
#| code-summary: R/simulate-chronology.R
#| eval: false
```

Example:

```{r}
#| label: example-chronology
set.seed(STARTREK)

chronology <- simulate_chronology(500, 3500, 5)
chronology
```

### Sample Dates

Following the generative model outlined in the main text, we sample years from 
the empirical distribution using inverse transform sampling and then use those 
years to sample radiocarbon determinations (measurements or ages) from the 
calibration curve.

This and the above `simulate_chronology()` function are based on 
`CreateDataVaryWeight()` used in (Heaton 2021, [github link](https://github.com/TJHeaton/NonparametricCalibration/blob/main/SimStudyFuncsFinal.R)).

```{r}
#| label: function-sample-dates
#| file: R/sample-dates.R
#| code-fold: true
#| code-summary: R/sample-dates.R
#| eval: false
```

Example:

```{r}
#| label: example-sample
set.seed(STARGATE)

samples <- sample_dates(chronology, 25, intcal20)
samples
```

### Calibrate Samples

Most of the algorithms outlined below rely on the calibrated dates as the main 
data for modeling.

```{r}
#| label: function-calibrate
#| file: R/calibrate.R
#| code-fold: true
#| code-summary: R/calibrate.R
#| eval: false
```

```{r}
#| label: calibrate-dates
calibrated_dates <- calibrate(samples, intcal20)
calibrated_dates
```

## Models

Here we specify our modeling algorithms for calibrating and summarizing 
radiocarbon dates. 

### Summed Probability Distribution

```{r}
#| label: spd
#| file: R/spd.R
#| code-fold: true
#| code-summary: R/spd.R
#| eval: false
```

Example:

```{r}
#| label: spd-example
#| time_it: true
spd(calibrated_dates)
```

### Composite Kernel Density

There are a variety of KDE methods for summarizing C14 samples (e.g. Bronk 
Ramsey 2017 and McLaughlin 2018), but we use the simplest possible approach 
here, just sampling from each sample's posterior and then running KDE on those 
samples, iterating this some number of times (500 by default). The final 
estimate of the true density is just the column means. 

```{r}
#| label: ckde
#| file: R/ckde.R
#| code-fold: true
#| code-summary: R/ckde.R
#| eval: false
```

Example:

```{r}
#| label: ckde-example
#| time_it: true
ckde(calibrated_dates)
```

### Bayesian Gaussian Mixture Model

This is roughly equivalent to the model proposed by Price et al (2021), but not
exactly the same. There are four notable changes. First, to specify the number 
of mixture components, we use kmeans with different values of K and select the 
one that minimizes BIC. Second, we use a small concentration for the Dirichlet 
prior to induce sparsity. Third, we let the prior on the variance adapt to the 
number of years in the interval. And fourth, to sample the posterior 
distribution, we implement adaptive dual averaging Hamiltonian Monte Carlo in a 
torch module, rather than using Stan.

```{r}
#| label: bgmm
#| file: R/bgmm.R
#| code-fold: true
#| code-summary: R/bgmm.R
#| eval: false
```

Example:

```{r}
#| label: bgmm-example
#| time_it: true
bgmm(calibrated_dates)
```

### Dirichlet Process Mixture Model

We use the DPMM with Polya Urn sampling as recommended by Heaton in the
[`carbondate` documentation](https://tjheaton.github.io/carbondate/reference/PolyaUrnBivarDirichlet.html):

> The library provides two slightly-different update schemes for the MCMC. In
> this particular function, updating of the DPMM is achieved by a Polya Urn
> approach (Neal 2000) This is our recommended updating approach based on
> testing.

Heaton also recommends an enormous number of iterations - the default is one
hundred thousand. This, however, is computationally prohibitive for any decently
large sample. That is why the simulation in (Heaton 2021) uses only 10,000
iterations, which we follow here.

```{r}
#| label: dpmm
#| file: R/dpmm.R
#| code-fold: true
#| code-summary: R/dpmm.R
#| eval: false
```

Example:

```{r}
#| label: dpmm-example
#| time_it: true
dpmm(samples)
```

### Gaussian Mixture Neural Network

NetC14 is a Gaussian Mixture Neural Network. To define that network, we first 
implement torch modules for the model's forward pass and custom loss. We then 
wrap these in a single model function that runs all the training logic, 
including back-propagation. 

```{r}
#| label: netc14
#| file: R/netc14.R
#| code-fold: true
#| code-summary: R/netc14.R
#| eval: false
```

Example:

```{r}
#| label: netc14-example
#| time_it: true
netc(calibrated_dates)
```

## Example

Here is a complete example of a single simulation.

```{r}
#| label: example
#| time_it: true
set.seed(STARTREK)

start <- 500
end <- 3500
n_clusters <- 5
n_samples <- 1250

chronology <- simulate_chronology(start, end, n_clusters)
samples <- sample_dates(chronology, n_samples, calibration = intcal20)
calibrated_dates <- calibrate(samples, intcal20)

models <- list(
  spd = spd(calibrated_dates),
  ckde = ckde(calibrated_dates),
  bgmm = bgmm(calibrated_dates),
  dpmm = dpmm(samples),
  netc = netc(calibrated_dates),
  oracle = chronology
)

densities <- lapply(models, density, years = seq(start, end))
```

```{r}
#| label: plot-example-densities
#| code-fold: true
#| output: false
svglite(
  filename = "figures/example-densities.svg",
  width = 5.5,
  height = 4
)

layout(
  matrix(c(1, 0, 0, 0, 0, 0, 2, 0, 3), nrow = 3, ncol = 3, byrow = TRUE),
  heights = c(0.7, 0.08, 1),
  widths = c(1, 0.01, 0.23)
)

par(defaults)
par(mai = c(0, 0.65, 0, 0))

plot(
  subset(intcal20, cal_age <= end & cal_age >= start),
  xlab = NULL,
  xaxs = "i",
  yaxs = "i"
)

par(mai = c(0.5, 0.65, 0, 0))

with(
  rbind(densities[["dpmm"]], densities[["netc"]]),
  new_plot(
    xlim = c(start, end),
    ylim = c(min(conf_lo, na.rm = TRUE), max(conf_hi, na.rm = TRUE)),
    xaxs = "i",
    yaxs = "i"
  )
)

x_axis()
y_axis()

x_title("Years Before Present (t)")
y_title("Density", line = 2.5)

add_grid()

with(densities, {
  plot(bgmm, add = TRUE, col = theme("BGMM"))
  plot(dpmm, add = TRUE, col = theme("DPMM"), se = TRUE)
  plot(netc, add = TRUE, col = theme("NetC14"), se = TRUE)
  plot(ckde, add = TRUE, col = theme("CKDE"))
  plot(spd, add = TRUE, col = theme("SPD"), lty = "dotted")
  plot(oracle, add = TRUE, col = theme("Oracle"), lwd = 2, lty = "dashed")
})

par(mai = rep(0, 4))
new_plot(c(0, 1), c(0, 1))
add_legend(
  "center",
  legend = c("True p(t)", "SPD", "CKDE", "BGMM", "DPMM", "NetC14"),
  col = theme("Oracle", "SPD", "CKDE", "BGMM", "DPMM", "NetC14"),
  text.col = theme("Oracle", "SPD", "CKDE", "BGMM", "DPMM", "NetC14"),
  lwd = rep(1.2, 6),
  lty = c("dashed", "dotted", "solid", "solid", "solid", "solid")
)

dev.off()
```

:::{#example-densities}
![](/figures/example-densities.svg){width=75%}
:::

Note: the above figure is pretty noisy, so we show confidence intervals only for 
the DPMM and NetC14.

**Estimate age**

Getting the posterior distribution over ages for each sample is just a matter
of multiplying the chronology across each row of the calibration matrix and then
dividing each row by its sum:

```{r}
#| label: example-posterior
v <- densities[["netc"]][["density"]]
q <- t(t(calibrated_dates) * v)
q <- q / rowSums(q)
```

The figure below demonstrates this. The blue line shows the calibration 
likelihood of a sample, the red dashed line shows the prior, i.e., the 
chronology or density estimated by the mixture, and the black line shows the 
posterior, or the product of the likelihood and the prior. The crucial result is 
to concentrate the probability mass at the nearest mode in the density, which is
presumably where the sample comes from.

```{r}
#| label: plot-example-age
#| code-fold: true
#| output: false
years <- seq(start, end)
curve <- subset(intcal20, cal_age >= start & cal_age <= end)

Si <- 2700

f14c_mu <- subset(curve, cal_age == Si)[["f14c_mu"]]
f14c_sigma <- f14c_mu * 25 / LIBBY

Ci <- calibrate_sample(f14c_mu, f14c_sigma, curve)
Ci <- Ci / sum(Ci)

modes <- c(2475, 2725)
v <- 0.2 * dnorm(years, modes[1], 40) + 0.8 * dnorm(years, modes[2], 40)
v <- v / sum(v)

Pi <- Ci * v
Pi <- Pi / sum(Pi)

support <- which(Ci > max(Ci) * 0.0001 | v > max(v) * 0.01)
xlim <- c(years[min(support)], years[max(support)])

svglite(
  filename = "figures/example-age.svg",
  width = 5.5,
  height = 4
)

layout(
  matrix(c(1, 0, 0, 0, 0, 0, 2, 0, 3), nrow = 3, ncol = 3, byrow = TRUE),
  heights = c(0.7, 0.08, 1),
  widths = c(1, 0.01, 0.23)
)

par(defaults)
par(mai = c(0, 0.65, 0, 0))

plot(
  subset(curve, cal_age <= xlim[2] & cal_age >= xlim[1]),
  xlab = NULL,
  xaxs = "i",
  yaxs = "i"
)

par(mai = c(0.5, 0.65, 0, 0))

new_plot(
  xlim = xlim,
  ylim = c(0, max(Ci, v, Pi) * 1.05),
  xaxs = "i",
  yaxs = "i"
)

x_axis()
y_axis()

x_title("Years Before Present (t)")
y_title("Density", line = 2.5)

add_grid()

abline(v = Si, col = coolors("gray"), lty = "dashed", lwd = 1.1)
text(
  x = Si + 10,
  y = max(Ci, v, Pi) * 1.03,
  labels = "True age",
  adj = c(0, 1),
  col = coolors("gray"),
  cex = 0.9
)

lines(years, Ci, col = coolors("blue"))
lines(years, v, col = coolors("orange"))
lines(years, Pi, col = coolors("brown"), lwd = 1.3)

par(mai = rep(0, 4))
new_plot(c(0, 1), c(0, 1))
add_legend(
  "center",
  legend = c("Likelihood", "Prior", "Posterior"),
  col = coolors("blue", "orange", "brown"),
  text.col = coolors("blue", "orange", "brown"),
  lwd = c(1, 1, 1.3),
  lty = c("solid", "solid", "solid")
)

dev.off()
```

:::{#example-age}
![](/figures/example-age.svg){width=75%}
:::

## Evaluation

We evaluate the models here against a large number of simulated chronologies and
radiocarbon samples. To save ourselves some time, we also run these simulations
in parallel. Generally speaking, running torch models in parallel is a bad idea 
since torch has built-in parallelism that could lead to thread contests. We try 
to account for this by limiting the number of threads in each parallel process 
to one. Running my PC at full tilt on 14 of 16 cores reduces the total wall time 
(in hours) for the evaluation by an order of magnitude.

Each simulation looks like the example above, following the generative model. 
The steps are as follows:

1. Sample a starting year.
2. Simulate a chronology.
3. Sample dates from the chronology.
4. Calibrate the dates.
5. Train models with dates.
6. Infer chronology and sample ages.
7. Evaluate models.

These steps are repeated for each simulation. Currently, there are 3 (n_years) * 
3 (n_samples) * 3 (n_clusters) = 27 starting conditions that are repeated 28 
times each, resulting in 756 unique simulations with randomization.

**Metrics**

```{r}
#| label: metrics
#| file: R/metrics.R
#| code-fold: true
#| code-summary: R/metrics.R
#| eval: false
```

**Simulations**

```{r}
#| label: model-evaluation
#| time_it: true
#| output: false
# define starting conditions for simulations
evaluation_grid <- expand.grid(
  n_years = c(500, 2500, 12500),
  n_samples = c(50, 250, 1250),
  n_clusters = c(1, 3, 9)
)

# repeat each combination n times
i <- rep(seq_len(nrow(evaluation_grid)), each = 28)
evaluation_grid <- evaluation_grid[i, ]
rownames(evaluation_grid) <- NULL

n_iter <- nrow(evaluation_grid)

# set up clusters
clusters <- makeCluster(14)

fns <- c(
  "bgmm.R",
  "calibrate.R",
  "ckde.R",
  "constants.R",
  "dataframe-helpers.R",
  "dpmm.R",
  "metrics.R",
  "mixture-helpers.R",
  "netc14.R",
  "sample-dates.R",
  "simulate-chronology.R",
  "spd.R"
)

clusterExport(
  clusters,
  varlist = c("fns", "evaluation_grid", "intcal20"),
  envir = environment()
)

clusterEvalQ(clusters, {
  library(carbondate)
  library(torch)
  torch_set_num_threads(1)
  lapply(file.path("R", fns), source)
})

# run simulations in parallel
evaluation_results <- parLapply(clusters, seq_len(n_iter), \(i) {
  set.seed(STARTREK + i)

  # scenario
  n_years <- evaluation_grid[["n_years"]][i]
  n_clusters <- evaluation_grid[["n_clusters"]][i]
  n_samples <- evaluation_grid[["n_samples"]][i]
  start <- sample(seq(100, 49500 - n_years), 1)
  end <- start + n_years - 1

  # simulate chronology, draw samples, and calibrate
  chronology <- simulate_chronology(start, end, n_clusters)
  samples <- sample_dates(chronology, n_samples, intcal20)
  calibrated_dates <- calibrate(samples, intcal20)

  # training
  models <- list(
    ckde(calibrated_dates),
    bgmm(calibrated_dates),
    dpmm(samples),
    netc(calibrated_dates),
    spd(calibrated_dates)
  )

  # evaluation
  years <- seq(start, end)
  true_ages <- samples[["cal_age"]]

  evaluations <- lapply(models, \(.model) {
    .density <- density(.model, years)[["density"]]
    .ages <- predict(.model, calibrated_dates)

    data.frame(
      tvd = tvd(.density, chronology),
      wd = wd(.density, chronology),
      mae = mae(.ages, true_ages),
      mse = mse(.ages, true_ages)
    )
  })

  evaluations <- do.call(rbind, evaluations)

  # baseline
  oracle_ages <- predict(chronology, calibrated_dates)
  mae_oracle <- mae(oracle_ages, true_ages)
  mse_oracle <- mse(oracle_ages, true_ages)

  # collect and return
  e_clusters <- vapply(models, count_components, integer(1))
  elapsed <- sapply(models, attr, which = "elapsed")

  sim <- data.frame(
    iter = i,
    model = c("CKDE", "BGMM", "DPMM", "NetC14", "SPD"),
    start = start,
    n_years = n_years,
    n_samples = n_samples,
    n_clusters = n_clusters,
    e_clusters = e_clusters,
    mae_oracle = mae_oracle,
    mse_oracle = mse_oracle,
    elapsed = elapsed
  )

  cbind(sim, evaluations)
})

# close parallel processes and collect results
stopCluster(clusters)
evaluation_results <- do.call(rbind, evaluation_results)
```

Save results.

```{r}
#| label: save-evaluation
write.csv(
  evaluation_results,
  "data/evaluation-results.csv",
  row.names = FALSE
)
```

## Results

```{r}
#| label: prepare-results
evaluation_results <- transform(
  evaluation_results,
  recovery = (mae - mae_oracle) / mae_oracle
)
```

These visualizations emphasize that radiocarbon dating is a multi-objective 
problem involving calibration (or age estimation) and summarization (or density
estimation) and that these objectives can sometimes trade-off each other, 
especially for mixture models where there is non-identifiability. The x-axis 
measures how well a model does at *summarizing* dates (at estimating the true 
density). TVD refers to "Total Variation Distance", which measures distance 
between two probability distributions. The y-axis measures how well a model does 
at *calibrating* dates (at estimating the true age of each sample). MAE is L1 
loss or "Mean Absolute Error".

Recovery refers to proportional improvement in calibration over an "oracle" 
model that knows the true age distribution or chronology.

$$
\rho = \frac{\text{Model} - \text{Oracle}}{\text{Oracle}}
$$

The idea behind recovery is to identify a theoretical limit on model 
performance, like even if you knew the true density, how much error would remain 
in trying to estimate ages given calibration curve and measurement uncertainty?

**Number of Samples**

Points are median values across simulations and sample sizes.

```{r}
#| label: plot-evaluate-n-samples
#| code-fold: true
#| output: false
xy <- aggregate(
  cbind(tvd, recovery) ~ model + n_samples,
  FUN = median,
  data = evaluation_results
)

svglite(
  filename = "figures/evaluate-objectives-nsamples.svg",
  width = 5.5,
  height = 3
)

xlim <- range(xy[["tvd"]])
ylim <- range(xy[["recovery"]])

layout(matrix(c(1, 0, 2), ncol = 3), widths = c(1, 0.01, 0.23))

par(defaults)
par(mai = c(0.5, 0.5, 0, 0))

new_plot(xlim, ylim)
add_grid()

for (mdl in unique(xy[["model"]])) {
  i <- which(xy[["model"]] == mdl)
  x <- xy[["tvd"]][i]
  y <- xy[["recovery"]][i]
  z <- xy[["n_samples"]][i]

  clr <- theme(mdl)
  cbg <- if (mdl == "SPD") "#FFFFFF" else theme(mdl, shade = -0.6)
  cex <- c("50" = 1.0, "250" = 1.3, "1250" = 1.7)[as.character(z)]

  lines(x, y, col = clr, lwd = 0.5)
  points(x, y, pch = 21, col = clr, bg = cbg, cex = cex)
}

x_axis()
y_axis()

x_title("SUMMARIZING [TVD]")
y_title("CALIBRATING [MAE-Recovery]")

par(mai = rep(0, 4))
new_plot(c(0, 1), c(0, 1))
lg <- add_legend(
  x = 0,
  y = 1,
  legend = c("BGMM", "CKDE", "DPMM", "NetC14", "SPD"),
  pt.bg = c(theme("BGMM", "CKDE", "DPMM", "NetC14", shade = -0.6), "#FFFFFF"),
  col = theme("BGMM", "CKDE", "DPMM", "NetC14", "SPD"),
  text.col = theme("BGMM", "CKDE", "DPMM", "NetC14", "SPD"),
  pch = 21,
  pt.cex = 1.1,
  lwd = 0.5
)

text(
  x = lg[["rect"]][["left"]] + 0.01,
  y = lg[["rect"]][["top"]] + 0.01,
  labels = "Models",
  adj = 0,
  col = coolors("gray")
)

lg <- add_legend(
  x = 0,
  y = 0.5,
  legend = c(50, 250, 1250),
  pt.bg = coolors("gray", shade = -0.6),
  col = coolors("gray"),
  text.col = coolors("gray"),
  pch = 21,
  pt.cex = c(1.0, 1.3, 1.7)
)

text(
  x = lg[["rect"]][["left"]] + 0.01,
  y = lg[["rect"]][["top"]] + 0.01,
  labels = "N Samples",
  adj = 0,
  col = coolors("gray")
)

dev.off()
```

:::{#evaluate-objectives-nsamples}
![](/figures/evaluate-objectives-nsamples.svg){width=75%}
:::

**Number of Clusters**

Points are median values across simulations and number of clusters.

```{r}
#| label: plot-evaluate-n-clusters
#| code-fold: true
#| output: false
xy <- aggregate(
  cbind(tvd, recovery) ~ model + n_clusters,
  FUN = median,
  data = evaluation_results
)

svglite(
  filename = "figures/evaluate-objectives-nclusters.svg",
  width = 5.5,
  height = 3
)

xlim <- range(xy[["tvd"]])
ylim <- range(xy[["recovery"]])

layout(matrix(c(1, 0, 2), ncol = 3), widths = c(1, 0.01, 0.23))

par(defaults)
par(mai = c(0.5, 0.5, 0, 0))

new_plot(xlim, ylim)
add_grid()

for (mdl in unique(xy[["model"]])) {
  i <- which(xy[["model"]] == mdl)
  x <- xy[["tvd"]][i]
  y <- xy[["recovery"]][i]
  z <- xy[["n_clusters"]][i]

  clr <- theme(mdl)
  cbg <- if (mdl == "SPD") "#FFFFFF" else theme(mdl, shade = -0.6)
  cex <- c("1" = 1.0, "3" = 1.3, "9" = 1.7)[as.character(z)]

  lines(x, y, col = clr, lwd = 0.5)
  points(x, y, pch = 21, col = clr, bg = cbg, cex = cex)
}

x_axis()
y_axis()

x_title("SUMMARIZING [TVD]")
y_title("CALIBRATING [MAE-Recovery]")

par(mai = rep(0, 4))
new_plot(c(0, 1), c(0, 1))
lg <- add_legend(
  x = 0,
  y = 1,
  legend = c("BGMM", "CKDE", "DPMM", "NetC14", "SPD"),
  pt.bg = c(theme("BGMM", "CKDE", "DPMM", "NetC14", shade = -0.6), "#FFFFFF"),
  col = theme("BGMM", "CKDE", "DPMM", "NetC14", "SPD"),
  text.col = theme("BGMM", "CKDE", "DPMM", "NetC14", "SPD"),
  pch = 21,
  pt.cex = 1.1,
  lwd = 0.5
)

text(
  x = lg[["rect"]][["left"]] + 0.01,
  y = lg[["rect"]][["top"]] + 0.01,
  labels = "Models",
  adj = 0,
  col = coolors("gray")
)

lg <- add_legend(
  x = 0,
  y = 0.5,
  legend = c(1, 3, 9),
  pt.bg = coolors("gray", shade = -0.6),
  col = coolors("gray"),
  text.col = coolors("gray"),
  pch = 21,
  pt.cex = c(1.0, 1.3, 1.7)
)

text(
  x = lg[["rect"]][["left"]] + 0.01,
  y = lg[["rect"]][["top"]] + 0.01,
  labels = "N Clusters",
  adj = 0,
  col = coolors("gray")
)

dev.off()
```

:::{#evaluate-objectives-nclusters}
![](/figures/evaluate-objectives-nclusters.svg){width=75%}
:::

## Real Data

Here we reproduce the empirical examples from Price et al (2021) and Heaton 
(2021). The first three (Raths, Iron Age, and Paleo) come from the Heaton paper. 
The last (Tikal) comes from the Price paper.

```{r}
#| label: radiocarbon-data
radiocarbon <- read.csv("data/radiocarbon.csv")

radiocarbon <- within(radiocarbon, {
  cal_age <- NA_integer_
  f14c_mu <- exp(-c14_age / LIBBY)
  f14c_sigma <- f14c_mu * c14_error / LIBBY
})

radiocarbon <- subset(
  radiocarbon,
  select = c(dataset, cal_age, f14c_mu, f14c_sigma)
)

class(radiocarbon) <- c("C14Samples", "data.frame")
```

```{r}
#| label: radiocarbon-data-examples
#| time_it: true
datasets <- unique(radiocarbon[["dataset"]])
results <- setNames(
  vector("list", length = length(datasets)),
  datasets
)

for (ds in datasets) {
  set.seed(STARTREK)

  samples <- subset(radiocarbon, dataset == ds)

  # add start and end attributes to samples (this includes 100-year buffer)
  i <- vapply(
    samples[["f14c_mu"]],
    \(.x) which.min(abs(.x - intcal20[["f14c_mu"]])),
    numeric(1)
  )
  cal_ages <- intcal20[["cal_age"]][i]
  time_range <- range(cal_ages) + c(-100, 100)

  # make sure it does not go past modern
  time_range <- pmax(0, time_range)

  start <- time_range[1]
  end <- time_range[2]

  attr(samples, "start") <- start
  attr(samples, "end") <- end

  calibrated_dates <- calibrate(samples, intcal20)

  models <- list(
    spd = spd(calibrated_dates),
    ckde = ckde(calibrated_dates),
    bgmm = bgmm(calibrated_dates),
    dpmm = dpmm(samples),
    netc = netc(calibrated_dates)
  )

  results[[ds]] <- lapply(models, density, years = seq(start, end))
}
```

```{r}
#| label: plot-real-radiocarbon
#| code-fold: true
#| output: false
for (i in seq_along(results)) {
  ds <- names(results)[i]
  densities <- results[[i]]

  start <- min(densities[[1]][["cal_age"]])
  end <- max(densities[[1]][["cal_age"]])

  svglite(
    filename = sprintf("figures/real-%s.svg", tolower(ds)),
    width = 5.5,
    height = 4
  )

  layout(
    matrix(c(1, 0, 0, 0, 0, 0, 2, 0, 3), nrow = 3, ncol = 3, byrow = TRUE),
    heights = c(0.7, 0.08, 1),
    widths = c(1, 0.01, 0.23)
  )

  par(defaults)
  par(mai = c(0, 0.65, 0, 0))

  plot(
    subset(intcal20, cal_age <= end & cal_age >= start),
    xlab = NULL,
    xaxs = "i",
    yaxs = "i"
  )

  par(mai = c(0.5, 0.65, 0, 0))

  with(do.call(rbind, densities), {
    if (ds == "Tikal") {
      ylim <- c(0, 0.0045)
    } else {
      ylim <- c(min(conf_lo, na.rm = TRUE), max(conf_hi, na.rm = TRUE))
    }

    new_plot(
      xlim = c(start, end),
      ylim = ylim,
      xaxs = "i",
      yaxs = "i"
    )
  })

  x_axis()
  y_axis()

  x_title("Years Before Present (t)")
  y_title("Density", line = 2.5)

  add_grid()

  with(densities, {
    plot(bgmm, add = TRUE, col = coolors("brown"), se = FALSE)
    plot(dpmm, add = TRUE, col = coolors("blue"), se = TRUE)
    plot(netc, add = TRUE, col = coolors("orange"), se = TRUE)
    plot(ckde, add = TRUE, col = coolors("yellow"), se = FALSE)
    plot(spd, add = TRUE, col = coolors("gray"), lty = "dotted", se = FALSE)
  })

  par(mai = rep(0, 4))
  new_plot(c(0, 1), c(0, 1))
  add_legend(
    "center",
    legend = c("SPD", "CKDE", "BGMM", "DPMM", "NetC14"),
    col = theme("SPD", "CKDE", "BGMM", "DPMM", "NetC14"),
    text.col = theme("SPD", "CKDE", "BGMM", "DPMM", "NetC14"),
    lwd = rep(1.2, 5),
    lty = c("dotted", "solid", "solid", "solid", "solid")
  )

  dev.off()
}
```

**Raths**

:::{#real-raths}
![](/figures/real-raths.svg){width=75%}
:::

**Iron Age**

:::{#real-iron}
![](/figures/real-iron.svg){width=75%}
:::

**PaleoIndians**

:::{#real-paleo}
![](/figures/real-paleo.svg){width=75%}
:::

**Tikal**

:::{#real-tikal}
![](/figures/real-tikal.svg){width=75%}
:::

## Session Info

```{r}
#| label: session-info
sessioninfo::session_info(pkgs = "attached")
```

## Appendix

A list of additional R functions used in this analysis and their definitions.

```{r}
#| label: function-dataframe-helpers
#| file: R/dataframe-helpers.R
#| code-fold: true
#| code-summary: R/dataframe-helpers.R
#| eval: false
```

```{r}
#| label: function-mixture-helpers
#| file: R/mixture-helpers.R
#| code-fold: true
#| code-summary: R/mixture-helpers.R
#| eval: false
```

```{r}
#| label: function-plot-calibration
#| file: R/plot-calibration.R
#| code-fold: true
#| code-summary: R/plot-calibration.R
#| eval: false
```

```{r}
#| label: function-plot-density
#| file: R/plot-density.R
#| code-fold: true
#| code-summary: R/plot-density.R
#| eval: false
```

```{r}
#| label: function-print
#| file: R/print.R
#| code-fold: true
#| code-summary: R/print.R
#| eval: false
```

```{r}
#| label: function-plot-helpers
#| file: R/plot-helpers.R
#| code-fold: true
#| code-summary: R/plot-helpers.R
#| eval: false
```

```{r}
#| label: function-plot-theme
#| file: R/plot-theme.R
#| code-fold: true
#| code-summary: R/plot-theme.R
#| eval: false
```