---
title: "Supplement B: Tuning NetC14"
date: today
published-title: "COMPILED"
format:
  html: 
    embed-resources: true
    fig-align: center
    fig-dpi: 300
    fig-responsive: true
    number-sections: true
    number-depth: 3
    toc: true
    toc-depth: 3
    smooth-scroll: true
    theme: cosmo
    highlight-style: a11y
    code-tools: true
    code-links:
      - text: kbvernon/netc14
        icon: github
        href: https://github.com/kbvernon/netc14
execute: 
  echo: true
  warning: false
  error: false
---

```{r}
#| label: time-code-chunks
#| echo: false
# we are deep into overkill territory here
knitr::knit_hooks$set(
  time_it = local({
    function(before, options) {
      if (before) {
        assign('.start_time', Sys.time(), envir = .GlobalEnv)
      } else {
        proc_time <- difftime(Sys.time(), .start_time, units = "secs")
        proc_time <- round(as.numeric(proc_time), 2)

        mins <- proc_time %/% 60
        secs <- proc_time %% 60

        proc_time <- if (mins > 0) {
          sprintf("%dm %.1fs", mins, secs)
        } else {
          sprintf("%.1fs", secs)
        }

        sprintf(
          "<p class='chunk-timing'>%s</p>",
          paste0("Code execution time: ", proc_time)
        )
      }
    }
  })
)
```

```{=html}
<style>
.code-copy-outer-scaffold:has(.chunk-timing) {
  position: relative;
}

.code-copy-outer-scaffold:has(.chunk-timing) .sourceCode.cell-code {
  margin-bottom: 2em;
}

.chunk-timing {
  position: absolute;
  bottom: -1.5em;
  right: 0;
  text-align: right;
  font-size: 0.75em;
  margin: 0;
  padding: 0;
  color: rgba(150, 150, 150, 1);
}

p:has(img) { 
  text-align: center; 
}
</style>

<script>
document.addEventListener("DOMContentLoaded", function() { 
  document.querySelectorAll(".chunk-timing").forEach(el => {
    const code = el.closest(".cell").querySelector(".code-copy-outer-scaffold");
    if (code) code.appendChild(el);    
  });
});  
</script>
```

## Introduction

**Goal:** Optimize NetC14 hyperparameters $\pi$ and $\gamma$ - sparsemax 
temperature and weight on precision penalty - with respect to model performance 
in (i) mean absolute error (MAE) of calibration (or error in estimating true 
ages of samples) and (ii) total variation distance of summary (or error in 
estimating true density from which samples are drawn).

**Data:** For model tuning, data are simulated according to the generative 
model.

**Method:** We sweep a fixed grid of ($\pi$, $\gamma$) configurations against a
fixed panel of simulated chronologies, scoring every configuration against the 
same set of chronologies to focus on variance across configurations. 

## R Preamble

```{r}
#| label: dependencies
#| output: false
library(parallel)
library(svglite)
library(torch)

fns <- c(
  "calibrate.R",
  "constants.R",
  "metrics.R",
  "mixture-helpers.R",
  "plot-helpers.R",
  "plot-theme.R",
  "print.R",
  "sample-dates.R",
  "simulate-chronology.R",
  "spd.R"
)

lapply(file.path("R", fns), source, verbose = FALSE)
```

**Constants**

```{r}
#| label: constants
#| file: R/constants.R
#| eval: false
```

## Data

```{r}
#| label: intcal20
intcal20 <- structure(
  read.csv("data/intcal20.csv"),
  class = c("Calibration", "data.frame"),
  curve = "IntCal20"
)

intcal20 <- transform(intcal20, f14c_mu = exp(-c14_age / LIBBY))
intcal20 <- transform(intcal20, f14c_sigma = f14c_mu * c14_error / LIBBY)
```

The tuning grid crosses $\pi \times \gamma$ configurations. The chronology grid 
crosses problem size, sample count, and mixture complexity scenarios. Every 
configuration is evaluated against every chronological scenario.

```{r}
#| label: tuning-grid
tuning_grid <- expand.grid(
  pi = seq(0.5, 2.0, by = 0.25),
  gamma = seq(0.5, 2.5, by = 0.25)
)

evaluation_grid <- expand.grid(
  n_years = c(500, 2500, 12500) - 1,
  n_samples = c(50, 250, 1250),
  n_clusters = c(1, 3, 9)
)

n_configs <- nrow(tuning_grid)
n_scenarios <- nrow(evaluation_grid)
```

## Tuning

To make this loop as efficient as possible, we precompute the chronologies and
draw samples from them.

```{r}
#| label: simulation-loop
#| output: false
#| time_it: true
simulations <- vector("list", n_scenarios)

for (i in seq_len(n_scenarios)) {
  set.seed(STARGATE + i)

  # scenario
  n_years <- evaluation_grid[["n_years"]][i]
  n_clusters <- evaluation_grid[["n_clusters"]][i]
  n_samples <- evaluation_grid[["n_samples"]][i]
  start <- sample(seq(100, 49500 - n_years), 1)
  end <- start + n_years

  # simulate chronology and draw samples
  chronology <- simulate_chronology(start, end, n_clusters)
  samples <- sample_dates(chronology, n_samples, intcal20)

  simulations[[i]] <- list(
    id = i,
    start = start,
    n_years = n_years,
    chronology = chronology,
    samples = samples
  )
}
```

Each chronology is simulated **once** and reused across all configurations. The 
configuration sweep only has to fit NetC14.

```{r}
#| label: tuning-loop
#| output: false
#| time_it: true
clusters <- makeCluster(14)

fns <- c(
  "bgmm.R",
  "calibrate.R",
  "constants.R",
  "dataframe-helpers.R",
  "metrics.R",
  "mixture-helpers.R",
  "netc14.R",
  "sample-dates.R",
  "simulate-chronology.R",
  "spd.R"
)

clusterExport(
  clusters,
  varlist = c(
    "fns",
    "intcal20",
    "n_configs",
    "tuning_grid"
  ),
  envir = environment()
)

clusterEvalQ(clusters, {
  library(torch)
  torch_set_num_threads(1)
  lapply(file.path("R", fns), source, verbose = FALSE)
})

tuning_results <- parLapply(clusters, simulations, \(scenario) {
  # scenario
  years <- with(scenario, seq(start, start + n_years - 1))
  calibrated_dates <- with(scenario, calibrate(samples, intcal20))
  true_ages <- scenario[["samples"]][["cal_age"]]
  true_density <- scenario[["chronology"]]

  # configurations
  results <- vector("list", n_configs)

  for (i in seq_len(n_configs)) {
    gamma <- tuning_grid[["gamma"]][i]
    pi <- tuning_grid[["pi"]][i]
    netc_mod <- netc(calibrated_dates, gamma = gamma, pi = pi)
    estimated_density <- density(netc_mod, years)[["density"]]
    estimated_ages <- predict(netc_mod, calibrated_dates)

    results[[i]] <- data.frame(
      pi = pi,
      gamma = gamma,
      tvd = tvd(estimated_density, true_density),
      wd = wd(estimated_density, true_density),
      mae = mae(estimated_ages, true_ages),
      mse = mse(estimated_ages, true_ages)
    )
  }

  # oracle
  oracle_ages <- predict(scenario[["chronology"]], calibrated_dates)
  maeo <- mae(oracle_ages, true_ages)
  mseo <- mse(oracle_ages, true_ages)

  # return
  results <- do.call(rbind, results)
  transform(
    results,
    scenario = scenario[["id"]],
    start = scenario[["start"]],
    maeo = maeo,
    mseo = mseo
  )
})

stopCluster(clusters)
tuning_results <- do.call(rbind, tuning_results)
```

Combine results and save.

```{r}
#| label: save-results
evaluation_grid[["scenario"]] <- seq_len(n_scenarios)
tuning_results <- merge(tuning_results, evaluation_grid, all.x = TRUE)

tuning_results <- tuning_results[c(
  "scenario",
  "pi",
  "gamma",
  "n_years",
  "n_samples",
  "n_clusters",
  "start",
  "tvd",
  "wd",
  "mae",
  "mse",
  "maeo",
  "mseo"
)]

write.csv(
  tuning_results,
  file = "data/tuning-results.csv",
  row.names = FALSE
)
```

## Results

The x-axis measures how well the model does at *summarizing* dates (at 
estimating the true density). TVD refers to "Total Variation Distance". The 
y-axis measures how well the model does at *calibrating* dates (at 
estimating the true age of each sample). MAE is L1 loss or "Mean Absolute 
Error". Note that we normalize the calibration score using an oracle model.

```{r}
#| label: aggregate-results
tuning_results <- transform(
  tuning_results,
  mae = (mae - maeo) / maeo
)

tuning_results <- aggregate(
  cbind(tvd, mae) ~ pi + gamma,
  data = tuning_results,
  FUN = median
)
```

```{r}
#| label: visualize-tuning-results
#| code-fold: true
#| output: false
svglite(
  "figures/tuning-objectives.svg",
  width = 5.5,
  height = 3
)

xlim <- quantile(tuning_results[["tvd"]], probs = c(0, 0.95), names = FALSE)
ylim <- quantile(tuning_results[["mae"]], probs = c(0, 0.95), names = FALSE)

par(defaults)
par(mai = c(0.5, 0.5, 0, 0))

new_plot(xlim, ylim)
add_grid()

with(
  tuning_results,
  points(
    tvd,
    mae,
    pch = 19,
    col = adjustcolor(coolors("gray"), 0.75),
    cex = 1
  )
)

x_axis()
y_axis()

x_title("SUMMARIZING [TVD]")
y_title("CALIBRATING [MAE-Recovery]")

dev.off()
```

:::{#tuning-objectives}
![](/figures/tuning-objectives.svg){width=75%}
:::

### Optimal configuration

The objectives of calibrating and summarizing radiocarbon dates should be 
complementary - a better estimate of the true density should lead to a better 
estimate of the true ages, and vice versa, but this is not guaranteed. 
Especially for mixture models, the issue of non-identifiability can lead to
models that do equally well at reconstructing the true density, but offer 
different point estimates of the true ages. This is most likely exacerbated by
uncertainty in the calibration curve.

The Pareto Front represents the set of configurations where you cannot improve 
one objective without sacrificing the other. Strictly speaking, we want the set 
of configurations that are non-dominated, meaning no other configuration beats 
them on both objectives.

```{r}
#| label: best-configuration
pareto <- function(x, y) {
  vapply(
    seq_len(length(x)),
    function(i) !any(x <= x[i] & y <= y[i] & (x < x[i] | y < y[i])),
    logical(1)
  )
}

subset(tuning_results, pareto(tvd, mae))
```

## Session Info

```{r}
#| label: session-info
sessioninfo::session_info(pkgs = "attached")
```